Power simulation section

### 2023-09-08 docs/manuscript.Rmd
Update manuscript.Rmd

<!-- General short intro paragraph, need for well-powered and analyzed experimental research -->
When planning experimental research it is essential to determine an appropriate sample size [@lakensSampleSizeJustification2022] to ensure that the results obtained are both robust and informative, and to use appropriate statistical models to analyze the data. However, the recent replication crisis in experimental research, such as in Human-Computer Interaction (HCI), has illustrated many challenges surrounding the reproducibility and reliability of findings [@robertsonImprovingStatisticalPractice2016]. As a results, there is a growing need for rigorous statistical methodology and the adoption of well-powered experimental designs. While accessible software solutions exists for simple statistical models and experimental designs, many experimental researchers lack the skills and tools to conduct power analyses for more complex research designs using the flexible generalized linear mixed models (GLMM) framework. In the present case study, we provide a tutorial comprised of a concrete example for custom a priori power analyses using data simulations based on GLMMs.

@@

<!-- Some basics on statistical power -->
In empirical research relying on hypothesis testing, the most common strategy for determining an adequate sample size is based on statistical power [@lakensSampleSizeJustification2022]. Statistical power is defined as the probability that a hypothesis test has a significant p-value when analyzing repeated samples from a population with a true effect of some pre-specified size. Less formally, power is also described as the probability that a hypothesis test accepts the alternative hypothesis $H_1$ if $H_1$ is indeed true. If the sample size (i.e., the number of participants and/or stimuli) used for data collection is insufficient to detect the effects with high probability the effects or relationships being investigated, this particular study would be considered "under-powered".

<!-- what are the resulting issues with under-powered studies? -->
Conducting under-powered research has many negative consequences. First, relying on under-powered experiments may yield inconclusive (if low power is taken into account) or misleading (if low power is ignored) results, hindering the accumulation of knowledge. Second, under-powered studies waste resources by consuming time, effort, and funding without delivering meaningful results.

@@

<!-- standard software is available for "simple" models and study designs -->
For simple statistical models, like t-tests, ANOVA, and linear regression, with common study designs (e.g., mean comparison between two groups) user-friendly software for power analysis is readily available [@faulStatisticalPowerAnalyses2009]. <!-- FP: we could reintroduce the old section her on: with more complex models and designs, power analysis is more complicated because no analytic solution is possible, power has to be estimated by simulation, and there is no user-friendly software that includes all relevant designs.; TK: yes let's do this. I am not entirely sure what you mean with "analytic" solution, i think we should define this.  -->

@@

<!-- intro into GLMMs -->
GLMMs are an extension of LMMs (Linear Mixed Models) which are in turn extensions of classic linear regression models that account for correlated data and hierarchical structures [@fahrmeirRegressionModelsMethods2021]. In this context, correlated data means that observations within a given dataset are not independent of each other. In other words, the values of the dependent variable for one observation may be related to the values of the dependent variable for other observations in a systematic way. This correlation can arise for various reasons, such as repeated measurements on the same subjects over time, observations clustered within certain groups or locations, or data collected at different levels of granularity. Hierarchical structures can be found in datasets that have a nested or hierarchical organization. For example, one might have data on students nested within schools, employees nested within departments, or patients nested within hospitals. In such cases, the observations within each higher-level grouping (e.g., school, department, hospital) tend to be more similar to each other than to observations in other groups.
LMMs are used when the outcome variable is continuous and follows a normal distribution (when conditioned on the predictor variables). They allow for the modelling of fixed effects, which capture the relationships between predictors and the outcome, as well as random effects, which account for the correlation and variability within groups or subjects. Random effects are typically assumed to follow a normal distribution with a mean of zero and a variance that quantifies the heterogeneity across the groups or subjects. 
As mentioned, GLMMs extend the LMM framework to accommodate non-normal and categorical outcome variables. They are used when the outcome variable does not follow a normal distribution, but instead belongs to a different distribution family (e.g., binomial). GLMMs incorporate both fixed and random effects, similar to LMMs, but also involve a link function that connects the linear predictor to the expected value of the outcome variable. The link function allows for modelling the relationship between predictors and the outcome in a non-linear way that is appropriate for the specific distribution family of the response variable.

@@

<!-- what are the existing option for a priori power analysis for GLMMs / where are existing tutorials limited -->
Currently available software packages for power simulations <!-- FP: because you removed the sentence on the difference between analytic and simulation based power analysis, it is now not clear why from now on we always talk about power simulation; TK: sorry if I have accidentally cut this distinction when merging versions, i think it is super important to have the definition between the two here.  --> do not include GLMMs, making it necessary to build data simulations tailored specifically to the study design [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022]. 
A number of tutorials have been published describing how to perform such power simulation for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021]. 
However, most of these tutorials only cover linear mixed models (LMMs) and focus on the most common designs (e.g., INSERT EXAMPLE). This narrow focus provides little guidance for researchers faced with more complex study designs (e.g., INSERT EXAMPLE), which are very common [@brownIntroductionLinearMixedEffects2021]. 
The necessary assumptions for power simulations in GLMMs include assumptions about the distributional form of the outcome variable, the random effects, and the error structure. The distributional assumption specifies the family of distributions for the outcome variable, such as Gaussian, Poisson, or binomial. Assumptions about the random effects include the assumption of normality and the covariance structure among the random effects. Additionally, assumptions about the error structure, such as independence or correlation, must be specified. Interpreting these assumptions entails understanding the underlying assumptions of the model and ensuring they align with the characteristics of the data being analyzed.
Here, existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, specifying assumptions a practical challenge [@maxwellSampleSizePlanning2008].

<!-- Summary / Wrap-Up, pick up main arguments and what the paper provides, keep this short! -->
Based on need for well-powered experimental research using GLMMs and the lack of tools to conduct corresponding power analyses, in this tutorial paper, we present a case study that serves as a practical demonstration of how to perform a data simulation and an a priori power simulation for a more complex study design. Thereby, we aim to equip researchers with the tools needed to simulate their own data and determine appropriate sample sizes for their own research.

@@

In this section, we outline the steps for performing data simulation and a priori power analysis for GLMMs, using a case study based on a specific experimental study design from the area of human-AI (artificial intelligence) interaction research.

@@

<!-- describe the study idea and design -->
In the present case study, we simulate the data for an experiment where the diagnostic performance of users of an AI-enabled diagnostic decision support system is to be evaluated. 
The goal is to understand how AI advice influences medical decision-making. Participants, radiologists (task experts) and students/interns (non-task experts) <!-- FP: during our work on the paper, I have heard two different stories of who the non experts are; either students/interns or full doctors that where not radiologists. Although this is only a data simulation example, it is nonetheless important here because the assumed effect sizes will probably differ! So who did you imagine when choosing the effect sizes?; SG can you clarify here please?  -->, review head computer tomography (CT) scans to assess the presence of bleeding, more specifically, intracranial haemorrhage (ICH). 
To support their decision-making, in two experimental conditions, an AI model provides initial diagnostic advice, which can be used as guidance by the participants. This AI advice can be either correct (80% of cases) or incorrect (20%). In the control condition, no AI advice is presented, meaning that the participants have to read the CT scan without any support. After reviewing the CT scan, participants deliver a medical diagnosis (bleeding or no bleeding), which may be either accurate or inaccurate.
This experimental design introduces some missing values by design since the advice is neither correct nor incorrect when no advice is present, which must be taken into account when simulating and analyzing the data. With this experiment, we want to determine if (a) experts are better than non-experts in reading head CT scans, (b) correct AI advice leads to better diagnostic accuracy than incorrect AI advice, and (c) different presentations of AI advice influence performance.
In this example, recruiting task experts (i.e., radiologists) is more challenging due to their limited availability, while non-experts (i.e., students/interns) are more readily accessible. 
The goal of the power simulation is to determine how many task experts and non-experts must be recruited to achieve sufficient statistical power in the planned experiment.

@@

<!-- deep dive into our specific example -->
In a GLMM, the expected value of the dependent variable Y conditioned on the vector of predictor variables $\mathbf{X}$ and random effects $\mathbf{U}$, transformed by a link function $g()$ is modeled as a linear combination $\eta$ of the predictor variables $\mathbf{X}$, the random effects $\mathbf{U}$ and the model parameters $\mathbf{\beta}$ [@fahrmeirRegressionModelsMethods2021]. $$
g(E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = \eta
$$ Equivalently, the conditional expected value is modeled as the linear combination $\eta$, transformed by the inverse link function $g^{-1}()$. $$
E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = g^{-1}(\eta)
$$ If the dependent variable (i.e., diagnostic accuracy) $Y$ is a binary variable with values $0$ or $1$, the conditional expected value is equivalent to the probability: $$
P_{si} := P(Y = 1|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})
$$ In our case study, $P_{si}$ is the conditional probability that a subject $s$ gives the correct response to item (i.e., CT scan) $i$.

@@

In the first six lines of the function definition, we set some default parameter values (which we will explain in a later section) and load the packages we use to manipulate and simulate data. 
In our case study, each subject (`n_subjects` in total) is assumed to respond to each item (i.e., CT scan; `n_items` in total). Thus the `add_random` command creates a fully-crossed `data.frame` with `n_subjects` $\times$ `n_items` rows. We add a between-subject effect with the `add_between` command, simulating that about $25\%$ of subjects are experts. The next two lines simulate that in about $\frac{2}{3}$ of trials, subjects will be presented with AI advice, and if advice is presented, the advice will be correct in about $80\%$ of cases (the variable `advice_correct` is always 0 when no advice is presented). Next, we simulate one random effect for each subject (`u0s`) and for each item (`u0i`). As assumed by standard GLMMs, the `add_ranef` function draws the random effects from a normal distribution with a mean 0 and a standard deviation specified by the user. With all design variables done, we are ready to simulate our model equation as outlined in equation X. The linear predictor variable `linpred` ($\eta$ in the GLMM model equations) combines the predictor variables, random effects and model parameters as assumed by our model. We then transform the linear predictor with the inverse-link function to compute `y_prob`, the probability that the subject correctly solved the item (in R, the inverse-logit link is computed with `plogis` and the logit link with `qlogis`). In the final step, we simulate the binary dependent variable `y_bin` (i.e., whether the subject chooses the correct diagnosis for the CT scan) by -- for each trial -- drawing from a Bernoulli distribution with success probability `y_prob`.

@@

Unfortunately, the model parameters in a binomial GLMM are hard to interpret because 1) the $\beta$ parameters are connected to the modeled probability via the non-linear inverse-logit link, and 2) we also have to consider the random effects. The most simple interpretation works by imagining a subject with average ability ($u_{0s} = 0$) responding to an item (i.e., CT scan) with average difficulty ($u_{0i} = 0$). Then the model implied probability that such a person solves such an item accurately is given by: $$
\begin{aligned}
P(Y=1|\mathbf{X=x}, \mathbf{U} = \mathbf{0}) = \\
= inverse\_logit[&\beta_0 + \beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
$$

@@

The `Fixed effects` table in Figure X also includes p-values for hypothesis tests with null hypotheses of the style $H_0: \beta = 0$. However, for many research questions of interest, we are not interested in these two-sided tests that refer to only a single parameter.

@@

Table X shows our set of assumptions concerning the probability that an average subject solves an average item for each experimental condition, as well as the corresponding equations implied by the model:

| *Experimental condition*  | $P(Y=1|\mathbf{X=x}, \mathbf{U} = \mathbf{0})$ | *Implied equation*                                                              |
|--------------------|:------------------------------:|--------------------|
| no advice, no expert      |                      0.70                      | $logit(0.70) = \beta_0$                                                         |
| no advice, expert         |                      0.90                      | $logit(0.90) = \beta_0 + \beta_e$                                               |
| false advice, no expert   |                      0.40                      | $logit(0.40) = \beta_0 + \beta_a$                                               |
| false advice, expert      |                      0.85                      | $logit(0.85) = \beta_0 + \beta_e + \beta_{a} + \beta_{ea}$                      |
| correct advice, no expert |                      0.90                      | $logit(0.90) = \beta_0 + \beta_a + \beta_c$                                     |
| correct advice, expert    |                      0.95                      | $logit(0.95) = \beta_0 + \beta_e + \beta_a + \beta_c + \beta_{ea} + \beta_{ec}$ |

@@

Although the derivations above are straightforward, it is important not to misinterpret their implications: In binomial GLMMs, the average probability to solve an item (averaged across persons of varying ability and items of varying difficulty) is **not** equal to the probability that a person with average ability solves an item with average difficulty. The first perspective implies a so-called marginal interpretation while the second one implies a conditional interpretation. For example, we determined the $\beta$ parameters in a way that corresponds to a desired conditional probability of $0.95$, that an expert with average ability solves an item with average difficulty when presented with correct advice. However, even if the model were true, we would not observe this probability value if we estimated the marginal probability in a group of experts responding to items presented with correct advice from a big sample of subjects drawn from their natural distribution of ability and items drawn from their natural distribution of difficulty.

The inequality of conditional and marginal effects in GLMMs [@fahrmeirRegressionModelsMethods2021] makes their interpretation more difficult. One must be careful when specifying parameter values based on previous studies or pilot data that use the marginal interpretation (e.g., a pilot study providing an estimate of how often neurologists make an accurate diagnosis based on brain scans). <!-- SG: What does that mean for someone who wants to do this then? FP: is it more clear now? --> However, this does not mean that we cannot use the marginal interpretation (average probability across persons and items) to inform plausible parameter values: When parameter values have been selected, we can compute the implied marginal distributions and compare this information to our domain knowledge. Then we can iteratively adjust the parameter values until we are satisfied with the implied distributions.

Earlier, we have already encountered one way to visualize the implied marginal distributions: We can fit our model to a simulated dataset and use the convenience functions from the marginaleffects package to compute averaged predictions that correspond to our quantities of interest. However, the model predictions will only be close to the true distribution if the simulated dataset is very large, but then the model fitting consumes a lot of time and memory. A more sophisticated strategy is to simulate a large dataset and directly compute the averages, contrasts and distributions we are interested in.

@@

Figure X shows the model implied marginal distributions, including the mean, 66% and 95% intervals. We can see that, indeed, the average probabilities (black dots) slightly differ from the probabilities of average subjects and items considered in the previous section. This difference increases with the variability of the random effects.

@@

The final plot demonstrates that these plots are also useful for spotting standard deviations that are specified too high. For this example, we have set $\sigma_S = 3$ and $\sigma_I = 3$. This implies that in each experimental condition, the probabilities that a subject solves an item are usually close to either 0 or 1, which is not a plausible assumption. For example, we would expect experts with low ability to solve a difficult item with a probability substantially larger than zero, when presented with wrong advice. <!-- SG: Maybe explain why this is not plausible. FP: like this?-->

@@

The result of the computation is a data.frame that contains the p-values of all tested hypotheses for each simulated dataset.
In some iterations (predominantly in conditions with small sample sizes), model estimation did not converge with the lme4 package. We do not remove these results because non-convergence can also happen when analyzing the real data we plan to collect, thus, we want to factor in this possibility to keep our simulation more realistic.

@@

Much has been written on the optimal amount of power to target in empirical research. The most prominent heuristic is to target a power of 0.8 (when combined with a type I error rate of $\alpha = 0.05$), but depending on the research goals of the study, there are often good reasons to move away from this standard [@lakensJustifyYourAlpha2018; @lakensSampleSizeJustification2022]. <!-- Maybe give an example?--> When target power has been specified, the number of subjects and the number of items in our study design can be traded against each other based on practical considerations. For the sake of the example, let the targeted power be indeed about 0.8, using an $\alpha$ of 0.05 to detect an effect of the expected size implied by our data simulation. This could be achieved by collecting data from 200 subjects (about 25% of which will be experts), each completing the same 50 items (with advice present in about 67% of cases which is correct in about 80% of cases with present advice). If collecting data from 200 subjects is not feasible, an alternative would be to recruit 150 subjects but increase the length of the experiment to over 70 items. However, 70 items might take too long to complete for the radiologists participating in the study, who have a busy schedule. The simulation suggests that it might also be possible to plan a shorter experiment with only 30 items if it is feasible to recruit an even higher number of subjects (\> 250, to be determined by additional rounds of power analysis). Design parameters that also affect power, and which could be investigated in the simulation to find a more optimal trade-off, are the ratio of experts, the frequency of whether advice is presented and whether it is correct.

>>>>

<!-- SG: In the intro power analysis, a priori power analysis and sample size calculation is now used interchangeably, should we use one term consistently?-->

<!-- General short intro paragraph, need for well-powered and analyzed experimental research -->
When planning experimental research, it is essential to determine an appropriate sample size [@lakensSampleSizeJustification2022] to ensure that the results obtained are both robust and informative, and to use appropriate statistical models to analyze the data. However, the recent replication crisis in experimental research, such as in Human-Computer Interaction (HCI), <!-- SG: is the HCI needed here? Also, can we claim a general replication crisis in experimental research? For instance, have no idea if experimental physics also has a problem. Maybe "in several disciplines --> has illustrated many challenges surrounding the reproducibility and reliability of findings [@robertsonImprovingStatisticalPractice2016]. As a result, there is a growing need for rigorous statistical methodology and the adoption of well-powered experimental designs. While software solutions exist for simple statistical models and experimental designs, many researchers lack the skills and tools to conduct power analyses for more complex research designs using the flexible generalized linear mixed models (GLMM) framework. In the present work, we provide a tutorial comprised of a concrete example for custom a priori power analyses using data simulations based on GLMMs.

@@

<!-- Some basics on statistical power -->
In empirical research relying on hypothesis testing, the most common strategy for determining an adequate sample size is based on statistical power [@lakensSampleSizeJustification2022]. Statistical power is defined as the probability that a hypothesis test has a significant p-value when analyzing repeated samples from a population with a true effect of some pre-specified size. Less formally, power is also described as the probability that a hypothesis test accepts the alternative hypothesis $H_1$ if $H_1$ is indeed true. If the sample size (i.e., the number of participants and/or stimuli) used for data collection is insufficient to detect the effects with high probability the effects or relationships being investigated <!-- SG: I think there is something wrong with the logic of this sentence. Should it be like this? If the sample size used for data collection is insufficient to detect the effects or relationships being investigated with high probability, the study would be considered "underpowered"-->, this particular study would be considered "underpowered".

<!-- what are the resulting issues with underpowered studies? -->
Conducting underpowered research has many negative consequences. First, relying on underpowered experiments may yield inconclusive (if low power is taken into account) or misleading (if low power is ignored) results, hindering the accumulation of knowledge. Second, underpowered studies waste resources by consuming time, effort, and funding without delivering meaningful results.

@@

<!-- standard software is available for "simple" models and study designs -->
For simple statistical models, like t-tests, ANOVA, and linear regression, with common study designs (e.g., mean comparison between two groups), user-friendly software for power analysis is readily available [@faulStatisticalPowerAnalyses2009]. <!-- FP: we could reintroduce the old section her on: with more complex models and designs, power analysis is more complicated because no analytic solution is possible, power has to be estimated by simulation, and there is no user-friendly software that includes all relevant designs.; TK: yes let's do this. I am not entirely sure what you mean with "analytic" solution, i think we should define this. SG: I am not sure what the plan is here. This sentence alone makes little sense. The argument that these tools cannot be used or adjusted for more complex models is missing. If you want to link it to the next section, it would argue that these simple statistical methods are often not sufficient for researchers to answer their research questions and that more complex designs are required. I would not go in detail here, that no tools for GLMM exist, since this argument is already mentioned below-->

@@

<!-- intro into GLMMs -->
GLMMs are an extension of LMMs (Linear Mixed Models), which are, in turn, extensions of classic linear regression models that account for correlated data and hierarchical structures [@fahrmeirRegressionModelsMethods2021]. In this context, correlated data means that observations within a given dataset are not independent of each other. In other words, the values of the dependent variable for one observation may be related to the values of the dependent variable for other observations in a systematic way. This correlation can arise for various reasons, such as repeated measurements on the same subjects over time, observations clustered within certain groups or locations, or data collected at different levels of granularity. Hierarchical structures can be found in datasets that have a nested or hierarchical organization. For example, one might have data on students nested within schools, employees nested within departments, or patients nested within hospitals. In such cases, the observations within each higher-level grouping (e.g., school, department, hospital) tend to be more similar to each other than to observations in other groups.
LMMs are used when the outcome variable is continuous and follows a normal distribution (when conditioned on the predictor variables). They allow for the modelling of fixed effects, which capture the relationships between predictors and the outcome, as well as random effects, which account for the correlation and variability within groups or subjects. Random effects are typically assumed to follow a normal distribution with a mean of zero and a variance that quantifies the heterogeneity across the groups or subjects. 
As mentioned, GLMMs extend the LMM framework to accommodate non-normal and categorical outcome variables. They are used when the outcome variable does not follow a normal distribution but instead belongs to a different distribution family (e.g., binomial) <!-- SG: do you think examples for the different types of variables are needed-->. GLMMs incorporate both fixed and random effects, similar to LMMs, but also involve a link function that connects the linear predictor to the expected value of the outcome variable. The link function allows for modelling the relationship between predictors and the outcome in a non-linear way that is appropriate for the specific distribution family of the response variable.

@@

<!-- what are the existing options for a priori power analysis for GLMMs / where are existing tutorials limited -->
Currently available software packages for power simulations <!-- FP: because you removed the sentence on the difference between analytic and simulation based power analysis, it is now not clear why from now on we always talk about power simulation; TK: sorry if I have accidentally cut this distinction when merging versions, i think it is super important to have the definition between the two here.  --> do not include GLMMs, making it necessary to build data simulations tailored specifically to the study design [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022]. 
A number of tutorials have been published describing how to perform such power simulation for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021]. 
However, most of these tutorials only cover linear mixed models (LMMs) and focus on the most common designs (e.g., INSERT EXAMPLE). This narrow focus provides little guidance for researchers faced with more complex study designs (e.g., INSERT EXAMPLE), which are very common [@brownIntroductionLinearMixedEffects2021]. 
The necessary assumptions for power simulations in GLMMs include assumptions about the distributional form of the outcome variable, the random effects, and the error structure. The distributional assumption specifies the family of distributions for the outcome variable, such as Gaussian, Poisson, or binomial. Assumptions about the random effects include the assumption of normality and the covariance structure among the random effects. Additionally, assumptions about the error structure, such as independence or correlation, must be specified. Interpreting these assumptions entails understanding the underlying assumptions of the model and ensuring they align with the characteristics of the data being analyzed.
Here, existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, specifying assumptions a practical challenge [@maxwellSampleSizePlanning2008].

<!-- Summary / Wrap-Up, pick up main arguments and what the paper provides, keep this short! -->
Based on the need for well-powered experimental research using GLMMs and the lack of tools to conduct corresponding power analyses, in this tutorial paper, we present a case study that serves as a practical demonstration of how to perform a data simulation and an a priori power simulation for a more complex study design. Thereby, we aim to equip researchers with the tools needed to simulate their own data and determine appropriate sample sizes for their own research.

@@

In this section, we outline the steps for performing data simulation and a priori power analysis for GLMMs using a case study based on a specific experimental study design from the area of human-AI (artificial intelligence) interaction research.

@@

<!-- describe the study idea and design -->
In the present case study, we simulate the data for an experiment where the diagnostic performance of users of an AI-enabled diagnostic decision support system is to be evaluated. 
The goal is to understand how AI advice influences medical decision-making. Participants, radiologists (task experts) and students/interns (non-task experts)<!-- FP: during our work on the paper, I have heard two different stories of who the non experts are; either students/interns or full doctors that where not radiologists. Although this is only a data simulation example, it is nonetheless important here because the assumed effect sizes will probably differ! So who did you imagine when choosing the effect sizes?; SG can you clarify here please? SG: Sure, when planning the study, our non-expert group was going to be students and interns. In a previous study, the non-expert groups were physicians with little task expertise. In the end, we did base some assumptions on these previous studies since we did not have anything else to work with -->, review head computer tomography (CT) scans to assess the presence of bleeding, more specifically, intracranial haemorrhage (ICH). 
To support their decision-making, in two experimental conditions <!--SG: I still think this will be confusing to readers. Why two experimental conditions?-->, an AI model provides initial diagnostic advice, which can be used as guidance by the participants. This AI advice can be either correct (80% of cases) or incorrect (20%). In the control condition, no AI advice is presented, meaning that the participants have to read the CT scan without any support. After reviewing the CT scan, participants deliver a medical diagnosis (bleeding or no bleeding), which may be either accurate or inaccurate.
This experimental design introduces some missing values by design since the advice is neither correct nor incorrect when no advice is present, which must be taken into account when simulating and analyzing the data. With this experiment, we want to determine if (a) experts are better than non-experts in reading head CT scans, (b) correct AI advice leads to better diagnostic accuracy than incorrect AI advice, and (c) different presentations of AI advice influence performance. <!--SG: c) has to be removed, if it is not in the analysis, right? -->
In this example, recruiting task experts (i.e., radiologists) is more challenging due to their limited availability, while non-experts (i.e., students/interns) are more readily accessible. 
The goal of the power simulation is to determine how many task experts and non-experts must be recruited to achieve sufficient statistical power in the planned experiment.

@@

<!-- deep dive into our specific example -->
In a GLMM, the expected value of the dependent variable Y conditioned on the vector of predictor variables $\mathbf{X}$ and random effects $\mathbf{U}$, transformed by a link function $g()$ is modeled as a linear combination $\eta$ of the predictor variables $\mathbf{X}$, the random effects $\mathbf{U}$ and the model parameters $\mathbf{\beta}$ [@fahrmeirRegressionModelsMethods2021]. $$
g(E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = \eta <!-- SG: Sorry for maybe a stupid question, but what does E stand for? and why are model parameters β mentioned but not in the equation?-->
$$ Equivalently, the conditional expected value is modelled as the linear combination $\eta$, transformed by the inverse link function $g^{-1}()$. $$
E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = g^{-1}(\eta)
$$ If the dependent variable (i.e., diagnostic decision)<!-- SG: not sure if diagnostic accuracy as a term is useful as a binary dependent variable since it is usually calculated as a percentage over many cases.)  --> $Y$ is a binary variable with values $0$ (i.e. inaccurate), or $1$ (i.e., accurate), the conditional expected value is equivalent to the probability: $$
P_{si} := P(Y = 1|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})
$$ In our case study, $P_{si}$ is the conditional probability that a subject $s$ gives the correct response to item (i.e., CT scan) $i$.

@@

In the first six lines of the function definition, we set some default parameter values (which we will explain in a later section) and load the packages we use to manipulate and simulate data. 
In our case study, each subject (`n_subjects` in total) is assumed to respond to each item (i.e., CT scan; `n_items` in total). Thus, the `add_random` command creates a fully-crossed `data.frame` with `n_subjects` $\times$ `n_items` rows. We add a between-subject effect with the `add_between` command, simulating that about $25\%$ of subjects are experts. The next two lines simulate that in $\frac{2}{3}$ of trials, subjects will be presented with AI advice, and if advice is presented, the advice will be correct in about $80\%$ of cases (the variable `advice_correct` is always 0 when no advice is presented). Next, we simulate one random effect for each subject (`u0s`) and for each item (`u0i`). As assumed by standard GLMMs, the `add_ranef` function draws the random effects from a normal distribution with a mean 0 and a standard deviation specified by the user. With all design variables done, we are ready to simulate our model equation as outlined in equation X. The linear predictor variable `linpred` ($\eta$ in the GLMM model equations) combines the predictor variables, random effects and model parameters as assumed by our model. We then transform the linear predictor with the inverse-link function to compute `y_prob`, the probability that the subject correctly solved the item (in R, the inverse-logit link is computed with `plogis` and the logit link with `qlogis`). In the final step, we simulate the binary dependent variable `y_bin` (i.e., whether the subject chooses the accurate diagnosis for the CT scan) by -- for each trial -- drawing from a Bernoulli distribution with success probability `y_prob`.

@@

Unfortunately, the model parameters in a binomial GLMM are hard to interpret because 1) the $\beta$ parameters are connected to the modelled probability via the non-linear inverse-logit link, and 2) we also have to consider the random effects. The most simple interpretation works by imagining a subject with average ability ($u_{0s} = 0$) responding to an item (i.e., CT scan) with average difficulty ($u_{0i} = 0$). Then the model implied probability that such a person solves such an item accurately is given by: $$
\begin{aligned}
P(Y=1|\mathbf{X=x}, \mathbf{U} = \mathbf{0}) = \\
= inverse\_logit[&\beta_0 + \beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
$$

@@

The `Fixed effects` table in Figure X also includes p-values for hypothesis tests with null hypotheses of the style $H_0: \beta = 0$. However, for many research questions of interest, we are not interested in these two-sided tests that refer to only a single parameter. <!-- Figure is not included in the pdf -->

@@

Table X shows our set of assumptions concerning the probability that an average subject solves an average item for each experimental condition, as well as the corresponding equations implied by the model:
<!-- SG Table number, title and note are missing. -->
| *Experimental condition*  | $P(Y=1|\mathbf{X=x}, \mathbf{U} = \mathbf{0})$ | *Implied equation*                                                              |
|--------------------|:------------------------------:|--------------------|
| no advice, no expert      |                      0.70                      | $logit(0.70) = \beta_0$                                                         |
| no advice, expert         |                      0.90                      | $logit(0.90) = \beta_0 + \beta_e$                                               |
| false advice, no expert   |                      0.40                      | $logit(0.40) = \beta_0 + \beta_a$                                               |
| false advice, expert      |                      0.85                      | $logit(0.85) = \beta_0 + \beta_e + \beta_{a} + \beta_{ea}$                      |
| correct advice, no expert |                      0.90                      | $logit(0.90) = \beta_0 + \beta_a + \beta_c$                                     |
| correct advice, expert    |                      0.95                      | $logit(0.95) = \beta_0 + \beta_e + \beta_a + \beta_c + \beta_{ea} + \beta_{ec}$ |

@@

Although the derivations above are straightforward, it is important not to misinterpret their implications: In binomial GLMMs, the average probability to solve an item (averaged across persons of varying ability and items of varying difficulty) is **not** equal to the probability that a person with average ability solves an item with average difficulty. The first perspective implies a so-called marginal interpretation, while the second one implies a conditional interpretation. For example, we determined the $\beta$ parameters in a way that corresponds to a desired conditional probability of $0.95$, that an expert with average ability solves an item with average difficulty when presented with correct advice. However, even if the model were true, we would not observe this probability value if we estimated the marginal probability in a group of experts responding to items presented with correct advice from a big sample of subjects drawn from their natural distribution of ability and items drawn from their natural distribution of difficulty.

The inequality of conditional and marginal effects in GLMMs [@fahrmeirRegressionModelsMethods2021] makes their interpretation more difficult. One must be careful when specifying parameter values based on previous studies or pilot data that use the marginal interpretation (e.g., a pilot study providing an estimate of how often neurologists make an accurate diagnosis based on brain scans). <!-- SG: What does that mean for someone who wants to do this then? FP: is it more clear now? SG: Definitely better; the question might still be what your solution to this issue is. --> However, this does not mean that we cannot use the marginal interpretation (average probability across persons and items) to inform plausible parameter values: When parameter values have been selected, we can compute the implied marginal distributions and compare this information to our domain knowledge. Then, we can iteratively adjust the parameter values until we are satisfied with the implied distributions.

Earlier, we have already encountered one way to visualize <!-- SG: so far, no figures in the pdf --> the implied marginal distributions: We can fit our model to a simulated dataset and use the convenience functions from the marginaleffects package to compute averaged predictions that correspond to our quantities of interest. However, the model predictions will only be close to the true distribution if the simulated dataset is very large, but then the model fitting consumes a lot of time and memory. A more sophisticated strategy is to simulate a large dataset and directly compute the averages, contrasts and distributions we are interested in.

@@

Figure X shows the model implied marginal distributions, including the mean, 66% and 95% intervals. We can see that, indeed, the average probabilities (black dots) slightly differ from the probabilities of average subjects and items considered in the previous section. This difference increases with the variability of the random effects. <!-- SG: Figure is not included in the PDF-->

@@

The final plot demonstrates <!-- SG: no figures in the pdf --> that these plots are also useful for spotting standard deviations that are specified too high. For this example, we have set $\sigma_S = 3$ and $\sigma_I = 3$. This implies that in each experimental condition, the probabilities that a subject solves an item are usually close to either 0 or 1, which is not a plausible assumption. For example, we would expect even experts with low ability to solve a difficult item with a probability substantially larger than zero even when presented with wrong advice. <!-- SG: Maybe explain why this is not plausible. FP: like this? SG: Better, but maybe still include a reason for this (years of training to do exactly this). Do you think it should be mentioned that already when just guessing, their probability would be .5-->

@@

The result of the computation is a data.frame that contains the p-values of all tested hypotheses for each simulated dataset.
In some iterations (predominantly in conditions with small sample sizes), model estimation did not converge with the lme4 package. <!-- SG: Maybe explain what this means--> We do not remove these results because non-convergence can also happen when analyzing the real data we plan to collect, thus, we want to factor in this possibility to keep our simulation more realistic.

@@

Much has been written on the optimal amount of power to target in empirical research. The most prominent heuristic is to target a power of 0.8 (when combined with a type I error rate of $\alpha = 0.05$), but depending on the research goals of the study, there are often good reasons to move away from this standard [@lakensJustifyYourAlpha2018; @lakensSampleSizeJustification2022]. <!-- Maybe give an example?--> When target power has been specified, the number of subjects and the number of items in our study design can be traded against each other based on practical considerations. For the sake of the example, let the targeted power be indeed about 0.8, using an $\alpha$ of 0.05 to detect an effect of the expected size implied by our data simulation. This could be achieved by collecting data from 200 subjects (about 25% of which will be experts), each completing the same 50 items (with advice present in about 67% of cases, which is correct in about 80% of cases with present advice). If collecting data from 200 subjects is not feasible, an alternative would be to recruit 150 subjects but increase the length of the experiment to over 70 items. However, 70 items might take too long to complete for the radiologists participating in the study, who have a busy schedule. The simulation suggests that it might also be possible to plan a shorter experiment with only 30 items if it is feasible to recruit an even higher number of subjects (\> 250, to be determined by additional rounds of power analysis). Design parameters that also affect power, and which could be investigated in the simulation to find a more optimal trade-off, are the ratio of experts, the frequency of whether advice is presented and whether it is correct.
